---
title: "Visualizing the Diamonds Data Set"
author: "Janna Mangasep"
date: '2022-08-26'
output: pdf_document
---

```{r setup, include=FALSE}
# Load dependencies
library(tidyverse) # 1.3.1
library(ggsci) # 2.9
library(statar) # 0.7.3
library(kableExtra, lib.loc = "/Library/Frameworks/R.framework/Versions/3.6/Resources/library") # 1.3.1
```

## Document Overview

For context, this R Markdown file revisits the work I previously did for my UChicago Harris Class "Data & Programming in R I." All work and outputs are my own and are simply guided by the questions from the assignment. This document aims to explore both variation and covariation in the `diamonds` data set, which is built-in from the `ggplot2` package within `tidyverse` and contains information on `r nrow(diamonds)` diamonds.

## Exploring Data Variation

### Price

To begin with, I examine the distribution of the `price` variable within `diamonds` through a histogram. Before doing so, I expected the distribution to be concentrated around the lower dollar prices. As a note, the maximum price in the data is rather extreme at `r max(diamonds$price)` dollars.

```{r pricevariation, echo=FALSE}
# Plot the variation of price with a histogram
ggplot(diamonds) +
  
  # Change the binwidth from the default 30 to 10 in order to look more closely
  ## at the data.
  geom_histogram(aes(x = price), binwidth = 10) +
  
  # Implement manually-calculated breaks for the x-axis tick marks, ensuring
  ## they're in intervals of $1,000.
  scale_x_continuous(
    breaks = seq(0, max(diamonds$price), 1000),
    labels = scales::dollar,
    minor_breaks = NULL
    ) +
  
  # Final touches (theme edits and informative labels)
  theme(
    text        = element_text(family = "serif"),
    plot.title  = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, vjust = 0.85)
    ) +
  labs(
    title = "Most diamonds are priced between $500 and $1,000",
    x     = "Price",
    y     = "Frequency"
    )
```

From the histogram above, there are a couple of things to note. First, most observations in `diamonds` have a `price` value between ~500 to ~1,000 dollars (i.e. it is *right-skewed*), and this aligns with my expectations. Also, there is a noticeable lack of observations around the 1,500 dollar mark, which is highly unusual as this is the only clear break in the distribution. 

### Carat

For the `carat` variable, I expect most of the observations to be concentrated around the lower values. This is because the previous exploration on `price` found most observations to be priced between 500 to 1,000 dollars. Therefore, assuming diamonds' prices are associated with their carats, most diamonds in the data are likely to have lower carats. 

Though my predictions for `carat` are similar to those for `price`, it is important to note that the possible values are very different between the two variables. While the latter ranges from `r min(diamonds$price)` dollars to `r max(diamonds$price)`, the former has a much smaller range of `r min(diamonds$carat)` carats to `r max(diamonds$carat)` carats. Therefore, I take precautions for proper scaling by winsorizing this variable (i.e., replacing carat outliers with specified quantiles). Please see [this documentation page](https://rdrr.io/cran/statar/man/winsorize.html) for more information on the `winsorize()` function from `statar`.

```{r caratvariation, echo=FALSE, message=FALSE}
# Create an updated column where the extreme values for carat are removed
diamonds_carat_reduced <- diamonds %>%
  mutate(carat_winsorized = statar::winsorise(carat, probs = c(0.01, 0.99)))

# Plot carats with winsorized values
ggplot(diamonds_carat_reduced) +
  
  # Reduce the binwidth size to look more closely at the distribution
  geom_histogram(aes(x = carat_winsorized), binwidth = 0.01) +
  
  # Implement manually-calculated breaks of 0.25 carats for the x-axis tick labels
  scale_x_continuous(breaks = seq(0.25, max(diamonds_carat_reduced$carat_winsorized), 0.25)) +
  
  # Include commas for y-axis values
  scale_y_continuous(labels = scales::comma) +
  
  # Final touches (theme edits and informative labels)
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(face = "italic", hjust = 0.5),
    text          = element_text(family = "serif")
    ) +
  labs(
    title    = "Most diamonds have around 0.3 carats, but there are multiple peaks in distribution",
    subtitle = "Carat values in data were winsorized by the 1st and 99th percentiles",
    x        = "Carat",
    y        = "Frequency"
    )
```

The most diamonds in the data appear to have around 0.3 carats. However, there are various, albeit smaller, peaks at other `carat` values (e.g., at 0.5 carats, ~0.7 carats, 1.0 carats, etc.). This does not fit with my expectations, as I expected this to be a smoother distribution, specifically a right-skewed distribution like with the variable `price`. Considering the strong concentration around *lower* prices (meaning from \$500 to \$1,000 dollars), this is a surprising find as there are variations within different ranges of carat values.

To better understand this distribution, I look into a sample of the peaks 0.5, 1.0, 1.5, and 2.0 within the entire `diamonds` data set (not winsorized), which I refer to as the "carat peaks". Specifically, I compare the frequency of diamonds at each peak and the frequency of diamonds at the carat value immediately before each peak (meaning 0.49, 0.99, etc.) to informally test whether sellers round up diamonds' carat values to increase prices.

```{r investigatecarat, echo=FALSE}
# Make function that summarizes the diamonds data based on given carat values
sum_carats <- function(carat_vector) {
  diamonds %>% 
    filter(carat %in% carat_vector) %>%
    count(carat, name = "Frequency")
}

# Create vector of sampled peak carat values
carat_peaks <- c(0.5, 1, 1.5, 2)

# Print summary tibble for carat peaks
kable(sum_carats(carat_peaks), caption = "Diamonds per Carat Peak") %>%
  kable_styling(latex_options = "HOLD_position")

# Print summary tibble for carats immediately before (0.01 carats) each peak
kable(sum_carats(carat_peaks - 0.01), caption = "Diamonds per Carat Immediately Before Peak") %>%
  kable_styling(latex_options = "HOLD_position")
```

From the tables above, it appears that significantly less diamonds have an "un-rounded" listed carat value. For example, only 23 diamonds are 0.99 carats, but 1,558 diamonds are 1 carat! Again, this discrepancy may come from diamond sellers rounding up the carat to sell diamonds for a greater profit. To investigate this further, I create the plot below to examine the relationship between `price` and `carat`

```{r pricecaratplot, echo=FALSE}
# Plot carat (winsorized) against price
ggplot(diamonds_carat_reduced, aes(x = carat_winsorized, y = price)) +
  
  # Plot smoothed line graph without region of standard error
  geom_smooth(se = FALSE, color = "black", size = 0.5) +
  
  # Show y-axis with dollar sign labels for style
  scale_y_continuous(labels = scales::dollar) +
  
  # Remove x-axis lines for style
  theme(
    panel.grid.major.x = element_blank(), 
    panel.grid.minor.x = element_blank()
    ) +
  
  # Final touches (theme edits and informative labels)
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(face = "italic", hjust = 0.5),
    text          = element_text(family = "serif")
    ) +
  labs(
    title    = "Diamonds with higher carats are associated with higher prices",
    subtitle = "Carat values in data were winsorized by the 1st and 99th percentiles",
    x        = "Carat",
    y        = "Price"
    )
```

From the plot above, it is apparent that there is a positive relationship between diamonds' carats and their listed prices. Therefore, it is reasonable to assume that diamond sellers would be incentivized to claim that a diamond is of higher carat than it is. This explains the dramatic peaks seen in the previous graph at rounded numbers. 




